---
title: "Multinomial and Ordinal Logistic Regression"
author: "Dr Irina Chis Ster"
date: "06/03/2021"
output:
  html_notebook:  
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# OUTLINE

-   Multinomial Logistic Regression - what is it and why do we need it?

-   Assumptions

-   Measures of associations and hypothesis testing

-   Goodness of fit

-   Model choice

-   Predictions

-   Ordinal Logistic Regression - what is it and why do we need it?

## Multinomial Logistic Regression

-   A(nother) generalized linear model

    -   linearity assumption extends from that between the outcome and
        independent variable (continuous outcome situation) to that
        between the log of the risk ratios and the potential explanatory
        variables

-   Supervised learning technique - parametric techniques

-   A regression technique applicable to nominal
    outcomes/responses/dependent variables with more than two categories

    -   Continuous outcomes - linear and multiple/multivariable
        regression

    -   Binary outcomes - univariate and multivariable/multiple logistic
        regression

    -   Nominal outcomes - multinomial logistic regression

-   Outcomes of different statistical types require different regression
    techniques

## Measures of associations

-   The outcome and the explanatory/independent variable(s)

-   Continuous outcome - [***regression coefficient***]{.ul} (NOT
    correlation)

-   Logistic regression - [***odds ratios (OR)***]{.ul}

-   Multinomial logistic regression - ([***relative) risk ratios
    (RRR)***]{.ul}

## Hypothesis testing (2-sided)

### Univariate association

-   [***Continuous outcome - linear regression***]{.ul}

    -   H0: no linear association between the outcome and the
        independent variable which translates into regression
        coefficient (slope of the regression line) $\beta=0$

    -   H1:there is a linear association between the outcome and the
        independent variable which translates into regression
        coefficient (slope of the regression line) $\beta\neq 0$

-   Binary outcome - Logistic regression

    -   H0: no association between the outcome and the independent
        variable translates into $OR=1$

    -   H1: there is an association between outcome and the independent
        variable which translates translates into $OR\neq 1$

-   Nominal data - Multinomial logistic regression

    -   H0: no association between outcome and independent variables
        translates into $RRR=1$

    -   H1: there is an association between outcome and the independent
        variable which translates into $RRR\neq 1$

### Multivariable associations

-   Continuous outcomes

    -   H0: All regression coefficients are 0

    -   H1: At least one is not 0

-   Logistic regression -

    -   H0: All OR=1

    -   H1: At least one OR not 1

-   Multinomial logistic regression -

    -   H0: All RRR=1

    -   H1: At least one RRR is not 1

## Assumptions

-   Independence of the observations

-   All models assume some kind of linear relationship:

    -   Continuous outcome: between the outcome and the independent
        variables

    -   Binary outcome: between the Log(OR) of the outcome and the
        independent variables

    -   Nominal outcomes: between the Log(RR) of the outcome and the
        independent variables

## Goodness of fit and model diagnosis

-   Continuous outcomes only: residuals play a central role in assessing
    model goodness of fit and diagnosis

-   Binary outcomes: no residuals. Hosmer-Lemeshow statistics

-   Categorical outcomes: the model is in essence a group of logistic
    regressions fit all at once to the data

    -   Similar results should hold if separate logistic regressions for
        binary outcomes constructed from the original nominal outcome

## Model choice

As for the all regression techniques:

-   Nested models - likelihood ratios comparisons

    -   H0: The two models are equally good

    -   H1: The two model with the larger number of explanatory
        variables is a better fit to the data

-   Non-nested models - AIC criterion: the smaller the value, the better
    the model. The AIC has no particular meaning per se.

## Interpretation:

-   Continuous outcome: a regression coefficient and its corresponding
    95%CI on either side of 0 indicates a statistically significant
    effect.

-   Categorical outcome: an RR (OR) and its corresponding 95%CI on
    either side of 1 indicates a statistically significant effect.

-   The magnitude and the sense of the effect differ according to the
    independent variable nature (continuous or categorical)

## Reminder:

-   Risk: $\pi=\frac{D}{N}$

-   Odds: $\frac{\pi}{1-\pi}=\frac{D}{\bar D}$

-   Risk Ratios: $\frac{\pi_{women}}{\pi_{men}}$

-   Odds Ratios: $\frac{OR_{women}}{OR_{men}}$

-   Relative Risk
    Ratios:$\frac{{\frac{RR_{women}}{RR_{men}}_{UK}}}{{\frac{RR_{women}}{RR_{men}}}_{France}}$

# Example 1: Social science data

-   US 1996 National Election Study: 10 variable subset of the 1996
    American National Election Study. Missing values and "don't know"
    responses have been list wise deleted. Respondents expressing a
    voting preference other than Clinton or Dole have been removed.
    (**Source:** Sapiro, Virginia, Steven J. Rosenstone, Donald R.
    Kinder, Warren E. Miller, and the National Election Studies.
    AMERICAN NATIONAL ELECTION STUDIES, 1992-1997: Ann Arbor, MI:
    University of Michigan)

-   Variables in the data:

    -   **Popul:** population of respondent's location in 1000s of
        people

    -   **TVnews** days in the past week spent watching news on TV

    -   **selfLR:** Left-Right self-placement of respondent: an ordered
        factor with levels

        -   extremely liberal (extLib)

        -   liberal (Lib)

        -   slightly liberal (sliLib)

        -   moderate (Mod)

        -   slightly conservative (sliCon)

        -   conservative (Con)

        -   extremely conservative (extCon)

    -   **ClinLR:** Left-Right placement of Bill Clinton (same scale as
        selfLR): an ordered factor with 7 levels as above

    -   **DoleLR:** Left-Right placement of Bob Dole (same scale as
        selfLR): an ordered factor with 7 levels as above

    -   **PID:** Party identification: an ordered factor with levels

        -   strong Democrat (strDem)

        -   weak Democrat (weakDem)

        -   independent Democrat (indDem)

        -   independent independent (indind)

        -   indepedent Republican (indRep)

        -   weak Republican (weakRep)

        -   strong Republican (strRep)

    -   **age:** Respondent's age in years

    -   **educ:** Respondent's education: an ordered factor with levels

        -   8 years or less (MS)

        -   high school dropout (HSdrop)

        -   high school diploma or GED (HS)

        -   some College (Coll)

        -   Community or junior College degree (CCdeg)

        -   BA degree (BAdeg)

        -   postgraduate degree (MAdeg)

    -   **Income**: Respondent's family income: an ordered factor with
        levels from less than 3K to 105Kplus in uneven i.e.
        (`$3K-$5K`,`$5K-$7K`,`$7K-$9K`,`$9K-$10K`,`$10K-$11K`,`$11K-$12K`,`$12K-$13K`,`$13K-$14K`,`$14K-$15K`,`$15K-$17K`,`$17K-$20K`,`$20K-$22K`,`$22K-$25K`,`$25K-$30K`,
        `$30K-$35`, `$35K-$40K`,
        `$40K-$45K`,`$45K-$50K, $50K-$60K`,`$60K-$75K`,`$75K-$90K`,
        `$90K-$105K`

    -   **Vote:** Expected vote in 1996 presidential election: a factor
        with level Clinton and Dole

## Reading the data

```{r}
mydata<-read.table("Survey.txt", header=T, sep="\t")
dim(mydata)
names(mydata)
```

## Creating a series of 3-category nominal variables

### The outcome

```{r}
mydata$PID_factor <- mydata$PID

## R order strings in alphabetical order
table(mydata$PID)

##create the new factor
levels(mydata$PID_factor) <- c("Independent","Independent", "Independent", "Democrat", "Republican", "Democrat","Republican")

table(mydata$PID_factor,mydata$PID)

```

### The education level

```{r}
mydata$educ_factor<-mydata$educ
table(mydata$educ)
levels(mydata$educ_factor) <- c("High","Average", "Average", "Average", "Low", "High","Low")
table(mydata$educ,mydata$educ_factor)

```

### The income level

```{r}
##this is a string which R orders in alphabetical order
table(mydata$income)

mydata$income_factor<-mydata$income
table(mydata$income_factor)
levels(mydata$income_factor) <- c("130", "5","7", "9", "10", "11", "12","13","14","15", "17", "20", "22", "25","3","30", "35", "40", "45","50","60", "75", "90", "105")
table(mydata$income_factor, mydata$income)

mydata$income_numeric<-as.numeric(mydata$income_factor)
is.numeric(mydata$income_numeric)
```

## The multinomial logistic - univariate model

### Continuous independent variable - summary at the log scale

```{r}
multinom_on_age <- multinom(mydata$PID_factor ~ age, data =mydata)
summary(multinom_on_age)
```

### Continuous independent variable - summary transformed for RR interpretation

```{r}
multinom_on_age_test <- summary(multinom_on_age)$coefficients/summary(multinom_on_age)$standard.errors

# 2-tailed z test
multinom_on_age_p <- (1 - pnorm(abs(multinom_on_age_test), 0, 1)) * 2
multinom_on_age_p

exp(coef(multinom_on_age))
exp(confint(multinom_on_age))

```

### Interpretation:

1.002= The relative risk ratio for a one year increase in **age** for
being Democrat relative to Independent. For one year increase in age,
the relative risk (probability) for being a Democrat relative to
Independent would be expected to increase by a factor of 1.002. The
p-value is 0.679 - so this is not statistically significant.

## The multinomial univariate model

### Categorical independent variable - summary at the log scale

```{r}
multinom_on_edu <- multinom(mydata$PID_factor ~ educ_factor, data =mydata)
summary(multinom_on_edu)

```

Interpretation for estimates at log scale:

-   The baseline of the outcome factor is "independent".

-   The baseline for education factor is "high".

-   There are 4 relative relationships:

    -   RR of Democrat vs. Independent for Average vs. High education

    -   RR of Democrat vs. Independent for Low vs. High education

    -   RR of Republic vs. Independent for Average vs. High education

    -   RR of Republic vs. Independent for Low vs. High education

```{r}
# 2-tailed z test
multinom_on_edu_test <- summary(multinom_on_edu)$coefficients/summary(multinom_on_edu)$standard.errors

multinom_on_edu_p <- (1 - pnorm(abs(multinom_on_edu_test), 0, 1)) * 2
multinom_on_edu_p
```

-   The p-values correspond to

    -   RR of Democrat vs. Independent for Average vs. High education
        (p=0.157)

    -   RR of Democrat vs. Independent for Low vs. High education
        (p=0.105)

    -   RR of Republic vs. Independent for Average vs. High education
        (p=0.895)

    -   RR of Republic vs. Independent for Low vs. High education
        (p=0.039)

### Categorical independent variable - summary transformed for RR interpretation

```{r}
exp(coef(multinom_on_edu))
```

-   There are 4 relative relationships:

    -   RR of Democrat vs. Independent is 1.28 higher for Average
        compared to High education

    -   RR of Democrat vs. Independent is 1.69 higher for Low compared
        to High education

    -   RR of Republican vs. Independent is 1.02 higher for Average
        compared to High education

    -   RR of Republican vs. Independent is 0.42 of Low compared to High
        education

```{r}
exp(confint(multinom_on_edu))

```

-   There are 4 relative 95%CI:

    -   RR of Democrat vs. Independent is 1.28 higher for Average
        compared to High education (0.91-1.81)

    -   RR of Democrat vs. Independent is 1.69 higher for Low compared
        to High education (0.897-3.167)

    -   RR of Republican vs. Independent is 1.02 higher for Average
        compared to High education (0.723 - 1.449)

    -   RR of Republican vs. Independent is 0.42 for Low compared to
        High education (0.184-0.958)

## The multinomial mutiple/multivariable model

### ADDITIVE MODEL

```{r}

multinom_on_additive <- multinom(mydata$PID_factor ~ income_numeric+ educ_factor, data =mydata)
summary(multinom_on_additive)

```

-   Interpretation -estimates at log scale

-   Same as above - but controlled/adjusted for income

-   In addition:

    -   RR of Democrat vs. Independent decreases with increasing
        income - controlled for education level

    -   RR of Republican vs. Independent increases with increasing
        income - controlled for education level

```{r}
multinom_on_additive_test <- summary(multinom_on_additive)$coefficients/summary(multinom_on_additive)$standard.errors

# 2-tailed z test
multinom_on_additive_p <- (1 - pnorm(abs(multinom_on_additive_test), 0, 1)) * 2
multinom_on_additive_p
```

-   The adjusted p-values correspond to

    -   RR of Democrat vs. Independent for Average vs. High education
        (p=0.162)

    -   RR of Democrat vs. Independent for Low vs. High education
        (p=0.136)

    -   RR of Republic vs. Independent for Average vs. High education
        (p=0.889)

    -   RR of Republic vs. Independent for Low vs. High education
        (p=0.042)

    -   RR of Democrat vs. Independent increase with income ( p=0.149)

    -   RR of Republican vs. Independent increase with income ( p=0.769)

```{r}
exp(coef(multinom_on_additive))
exp(confint(multinom_on_additive))

```

-   There are 6 relative 95%CI:

    -   Adjusted RR of Democrat vs. Independent is 1.28 higher for
        Average compared to High education (0.91-1.81) - so the RRR=1.28

    -   Adjusted RR of Democrat vs. Independent is 1.61 higher for Low
        compared to High education (0.859-3.052)- so the RRR=1.61

    -   Adjusted RR of Republican vs. Independent is 1.02 higher for
        Average compared to High education (0.723 - 1.451)

    -   Adjusted RR of Republican vs. Independent is 0.42 for Low
        compared to High education (0.186-0.969)

    -   Adjusted change in RR of Democrat vs. Independent for one unit
        increase in income (drops by (1-0.957)\*100% with one unit
        increasing in outcome)

    -   Adjusted change in RR of Republican vs. Independent for one unit
        increase in income(drops with increasing outcome)

## The multinomial mutiple/multivariable model

### INTERACTION MODEL

Are the effects of education on the RR of Democrat/Republican vs.
Independent modified by the income level? Or, the other way around, is
the effect of income level on the RR of Democrat/Republican vs.
Independent modified by the education level?

Fitting the model

```{r}

multinom_on_multiplic <- multinom(mydata$PID_factor ~ income_numeric* educ_factor, data =mydata)
summary(multinom_on_multiplic)
```

The p-values

```{r}
# 2-tailed z test
multinom_on_multiplic_test <- summary(multinom_on_multiplic)$coefficients/summary(multinom_on_multiplic)$standard.errors

multinom_on_multiplic_p <- (1 - pnorm(abs(multinom_on_multiplic_test), 0, 1)) * 2
multinom_on_multiplic_p
```

The RR

```{r}
exp(coef(multinom_on_multiplic))
exp(confint(multinom_on_multiplic))
```

#### Goodness of fit

```{r}
logitgof(mydata$PID_factor, fitted(multinom_on_multiplic)  )

```

#### Model choice

```{r}
lrtest(multinom_on_multiplic, multinom_on_additive)
AIC(multinom_on_multiplic, multinom_on_additive)
```

### PREDICTIONS

#### Create a speadsheet of values within the data range

##### Predictions by education factor at the means of income levels

```{r}
create_data_pred <- data.frame(educ_factor = c("Low", "Average", "High"), income_numeric = mean(mydata$income_numeric))
predict(multinom_on_multiplic, newdata = create_data_pred, "probs")
```

[***Interpretation:***]{.ul} these are the absolute proportions (risk)
of being across each education level and for
independent/democrat/republican at the average income. The proportions
add to 1 across every row (education level).

##### Predictions by education factor at detailed by income levels

```{r}
create_data_pred_2 <- data.frame(educ_factor = rep(c("Low", "Average", "High"), each = 128), income_numeric = rep(c(3:130),1))

## store the predicted probabilities for each value of education and income
pp.income <- cbind(create_data_pred_2, predict(multinom_on_multiplic, newdata =create_data_pred_2, type = "probs", se = TRUE))

## calculate the mean probabilities within each level of education
by(pp.income[, 3:5], pp.income$income_numeric, colMeans)
```

```{r}
## melt data set to long for ggplot2
lpp <- melt(pp.income, id.vars = c("educ_factor", "income_numeric"), value.name="probability")
head(lpp)  # view first few rows
```

```{r}
## plot predicted probabilities across income values for each level of education

ggplot(lpp, aes(x = income_numeric, y = value, colour = educ_factor)) + geom_line() + facet_grid(variable ~., scales = "free")
```

# Ordinal logistic regression

-   Applied to an ordinal dependent (response) variables

-   The independent variables may be categorical factors or continuous
    covariates.

-   One of the assumptions underlying ordered logistic regression is
    that the relationship between each pair of outcome groups is the
    same.

-   In other words, ordered logistic regression assumes that the
    coefficients that describe the relationship between, say, the lowest
    versus all higher categories of the response variable are the same
    as those that describe the relationship between the next lowest
    category and all higher categories, etc.

-   This is called the [***proportional odds assumption or the parallel
    regression assumption***]{.ul}.  This can be checked by [***Brant
    test***]{.ul} in R:

    -   H0: the proportional odds assumption holds

    -   H1: the proportional odds assumption does not hold

-   If [***Brant test***]{.ul} indicates that the parallel regression
    assumption is violated, we would need different models to describe
    the relationship between each pair of outcome groups.

-   A midway between regression on continuous outcomes and nominal
    outcomes

-   Model choice - similar in all regression techniques.

# Example 2: Social science

-   Cross sectional study in a population of young college attendants
    asked how likely is that they would apply to study for further
    university degrees

-   [***Outcome/Dependent/Response variable:***]{.ul} a three level
    variable called **apply** with levels "unlikely", "somewhat likely",
    and "very likely", coded 1, 2, and 3, respectively.

-   Three [***independent variables***]{.ul}::

    -   **pared**: a binary 0/1 variable indicating whether at least one
        parent has a graduate degree;

    -   **public**: a binary 0/1 variable indicating that the
        undergraduate institution is public (1) and (0) private, and

    -   **gpa**: which is the student's grade point average - continuous

## Reading and exploring the data

```{r}
mydata<-read.table("OrdinalLogisticExample.txt", header=T, sep="\t")
mydata
names(mydata)

##this is an ordinal outcome as we cannot mathematically operate in any way

## one at a time, table apply, pared, and public
lapply(mydata[, c("apply", "pared", "public")], table)
## three way cross tabs (xtabs) and flatten the table
ftable(xtabs(~ public + apply + pared, data = mydata))
summary(mydata$gpa)
sd(mydata$gpa)

##explores the outcome by the three independent variables
##We can also examine the distribution of gpa at every level of applyand broken down by public and pared. This creates a 2 x 2 grid with a boxplot of gpa for every level of apply, ##for particular values of pared and public. 

##To better see the data, we also add the raw data points on top of the box plots, with a small amount of noise (often called "jitter") and 50% transparency so they do not overwhelm the boxplots.

##Finally, in addition to the cells, we plot all of the marginal relationships. The margins make the final plot a 3 x 3 grid. In the lower right hand corner, is the overall relationship between apply and gpa.  

ggplot(mydata, aes(x = apply, y = gpa)) +
  geom_boxplot(size = .75) +
  geom_jitter(alpha = .5) +
  facet_grid(pared ~ public, margins = TRUE) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

## Creating an ordinal outcome

```{r}
mydata$apply_factor<-as.factor(mydata$apply_cat)
table(mydata$apply_factor,mydata$apply)

```

## Fitting a multivariable ordinal logistic regression

```{r}
## fit ordered logit model and store results 'model1'
model1 <- polr(apply_factor ~ public* gpa+ pared  , data = mydata, Hess=TRUE)

## view a summary of the model
summary(model1)

```

## The associated p-values

```{r}
(ctable <- coef(summary(model1)))
## calculate and store p values
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
 
## combined table
(ctable <- cbind(ctable, "p value" = p))

```

## The ODDS RATIOS

```{r}
## ci
(ci <- confint(model1)) # default method gives profiled CIs

## odds ratios
exp(coef(model1))

## OR and CI
exp(cbind(OR = coef(model1), ci))
```

## Interpretation for the additive models

0.94 = this is the proportional odds ratio of comparing public to
private on **apply** given the other variables in the model are held
constant.

For public school participants, the odds of very likely (to apply)
versus the combined somewhat likely and unlikely (to apply) are 0.94
times lower than for private school pupils, given the other variables
are held constant.

Likewise, the odds of the combined very likely and somewhat likely (to
apply) versus the unlikely (to apply) for public school participants are
0.94 times lower than for private school pupils, given the other
variables are held constant.

```{r}
## only brant model is validated as proportional assumptions
brant(model1)
```

There is no evidence to suggest that the model does violate the parallel
assumption.

```{r}
newdat <- data.frame(
  pared = rep(0:1, 200),
  public = rep(0:1, each = 200),
  gpa = rep(seq(from = 1.9, to = 4, length.out = 100), 4))

newdat <- cbind(newdat, predict(model1, newdat, type = "probs"))

##show first few rows
head(newdat)

lnewdat <- melt(newdat, id.vars = c("pared", "public", "gpa"), variable.name = "Level", value.name="Probability")
dim(lnewdat)
## view first few rows
lnewdat
table(lnewdat$public) 

ggplot(lnewdat, aes(x = gpa, y =value, colour = variable)) +
  geom_line() + facet_grid(pared ~ public, labeller="label_both")
```
